EBASE manuscript

### 2023-03-29 manu-draft.qmd
almost done with rgn edits

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "ebaseiotab")`).

@@

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "ebaseiotab")`. Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.

@@

<!-- edit -->
```{r}
# percent negative nse not do
nseall <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  filter(var != 'DO_mod') %>% 
  na.omit %>% 
  pull(nse)
perneg <- (sum(nseall < 0) / length(nseall) * 100) %>% 
  round() %>% 
  as.character() %>% 
  paste0('%')

@@

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameters estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.  However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).  These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model.  Results using EBASE default priors (medium values for $\mu$, high values for $\rho$) showed that the $b$ parameter can be estimated with reasonable accuracy, although still with considerable variability.  Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).  The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\rho$.

### EBASE comparison with real observations

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASE, and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08. For better comparison with the EBASE methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21]. The input data were provided to each method after converting the units as required (e.g., BASE requires PAR as $\mu$mol/m$^{2}$/s). Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth (i.e., $\text{mmol}/\text{m}^3/\text{d}$). Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection. All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASE was informed by results from the Odum output. Initial assessments indicated that unstable and inaccurate results for the time series are obtained from BASE using an uninformed prior distribution for $K$. All other prior distributions for BASE and EBASE were set as the software default. A one day model optimization period was used for EBASE for comparability with BASE.

The common set of metabolic parameters returned by the Odum, BASE, and EBASE methods are net ecosystem metabolism, gross production, respiration, and gas exchange. Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher values). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.

@fig-apacmpfig and `r run_reference(id = "apacmptab")` provide a comparison of the EBASE metabolic estimates with those from the Odum and BASE methods. Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The only exception was the comparison of respiration for the Odum and EBASE methods ($\rho$ = 0.11), where the former produced a much larger range of estimates. The intercept values were all significant, suggesting some level of bias when comparing estimates between the methods. Slope estimates were all significantly different from one for the comparisons using the observed dissolved oxygen time series as input, whereas only three of the eight comparisons were significantly different using the detided time series. This latter comparison suggests that detiding the time series produces more comparable estimates between methods. Similarly, the standard deviation ($\rho$) of the residuals of each regression comparison were generally lower in the detided results for all comparisons. Variation in the gas exchange estimates were also more similar comparing the Odum and EBASE methods that used the same formulation [@Wanninkhof14] regardless of input time series.

As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE and BASE are more subtle as anomalous values cannot be returned based on constraints in the mass-balance equations (e.g., it is impossible for respiration to be positive). However, the effects of tidal advection produced a wider range of estimates for NEM in EBASE and BASE compared to the Odum method ([@fig-apacmpfig]a), caused by greater variation in the production and respiration estimates. Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "apacmptab")`), regardless of the method used.

@@

![The default prior distributions for the $a$, $r$, and $b$ parameters used in EBASE. All priors are normal distributions and truncated to positive values. The $b$ parameter is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}

@@

![Comparisons of daily metabolic estimates from the Odum, BASE, and EBASE methods for one year of continuous monitoring data at Apalachicola Bay. Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model. All values are volumetric as mmol O$_2$ m$^3$ d$^{-1}$. Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}

@@

```{r}
#| label: ebaseiotab
#| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output.  The model notation for production, respiration, and gas exchange shows both the simple parameter name and the full model component used to estimate the parameter."
ebaseiotab
```

@@

```{r}
#| label: apacmptab
#| tbl-cap: "Summary statisics of daily metabolic estimates comparing Odum and BASE to EBASE for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D).  Separate comparisons are made for estimates based on observed and detided dissolved oxygen.  Summary statistics include Pearson correlation coefficients and intercept, slope, and residual standard deviation of a linear fit between the same metabolic estimates obtained from each method. Significance of the p-values for the intercept and slope indicates if the estimate is different from zero or one, respectively."
apacmptab
```

>>>>

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($dC$) per unit time ($dt$) of the input data. Production is modeled linearly as a function of surface irradiance ($PAR$, Watts m$^{-2}$) and $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$). Respiration is modeled with the single parameter $r$ ($\text{mmol}/\text{m}^3/\text{d}$). Gas exchange is modeled using the formulation in @Wanninkhof14 as a wind-based parameterization for the gas transfer velocity [@Ho16] multiplied by the oxygen concentration gradient as the difference between the concentration measured in the surface water layer, $C$ (mmol m$^{-3}$), and the saturation (or equilibrium) concentration, $C_{Sat}$ (mmol m$^{-3}$), calculated from salinity (psu) and temperature ($^\circ$C) [@Millero81; @Garcia92]. The gas transfer velocity is modeled using wind speed squared at 10 meters above the water ($U_{10}^2$, m$^2$ s$^{-2}$), water column depth ($H$ in meters, as a fixed parameter or time series from a pressure sensor), and the Schmidt number ($Sc$) defined as the ratio of the kinematic viscosity of water to the molecular diffusivity of the gas for oxygen [from water temperature and salinity using the polynomial fit in @Wanninkhof14]. The $b$ parameter is estimated from the data and has units of $\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$. As such, the three parameters estimated by EBASE are $a$, $r$, and $b$ using the required input data that includes time series of dissolved oxygen, water temperature, salinity, PAR, wind speed, and water column depth (`r run_reference(id = "tbl-ebaseiotab")`).

@@

The remaining three parameters $a$ (light efficiency, $\left(\text{mmol}/\text{m}^2/\text{d}\right)/\left(\text{W}/\text{m}^2\right)$), $r$ (respiration, $\text{mmol}/\text{m}^3/\text{d}$), and $b$ ($\left(\text{cm}/\text{hr}\right)/\left(\text{m}^2/\text{s}^2\right)$) are estimated by likelihood given the observed data and prior distributions for the parameters. The JAGS software [as for BASE, @Plummer03] is used with the EBASE R package [described below, @Beck22; @RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses based on the prior distributions, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps. The estimated dissolved oxygen concentration is also returned, which can be compared to the observed values as one measure of model performance. All parameter inputs and outputs for EBASE are shown in `r run_reference(id = "tbl-ebaseiotab")`. Areal rates for $P$, $R$, and $D$ (mmol m$^-2$ d$^{-1}$) are returned for comparability of estimates across locations where water column depth may vary.

@@

```{r}
# percent negative nse not do
nseall <- apasumdat %>% 
  tidyr::unnest('ests') %>% 
  filter(var != 'DO_mod') %>% 
  na.omit %>% 
  pull(nse)
perneg <- (sum(nseall < 0) / length(nseall) * 100) %>% 
  round() %>% 
  as.character() %>% 
  paste0('%')

@@

The actual model results for the best and worst performing combinations of prior distributions for the one day, seven day, and thirty day optimization periods are shown in @fig-optex (all results are averaged at the time-step of the optimization period for comparability). The best performing models were those with the highest median NSE values across all parameter estimates in @fig-priorcomp. The best performing models for each optimization period (@fig-optex a, c, e) showed similar results where the parameters estimates were similar to the synthetic time series, although more variation in the EBASE results were of course observed with the one day optimization period.  However, the $b$ parameter was estimated with high variability for the one and seven day optimization periods (@fig-optex a, c, bottom row) and with a large positive bias for the 30 day optimization period (@fig-optex e, bottom row).  These results were not unexpected because it was impossible to evaluate NSE estimates using a fixed value in the synthetic results and the accuracy was not considered in identifying the best performing model. Finally, the worst performing model for each optimization period (@fig-optex b, d, f) demonstrates the negative NSE values that were obtained (i.e., EBASE values much lower or higher than the synthetic values).  The worst performing models were caused by constraining EBASE to a small parameter space using low or high values for $\mu$ combined with low values for $\rho$.

### EBASE and Odum comparison with real observations

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The observed one-year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that uses a gas-exchange approach described in @Thebault08. For better comparison with the EBASE methods, the gas exchange parameterization in the Odum method was modified to use the @Wanninkhof14 approach, the the $b$ parameter fixed at 0.251 $\buni$ [as available in the WtRegDO R package, @Beck21]. An additional and fundamental difference between EBASE and the Odum method is the former is likelihood based with metabolic rates estimated by fitting the model to the data, whereas the latter is strictly arithmetic based primarily on integration of the diel dissolved oxygen curve. Thus, potentially different results are expected between the approaches based on differences in formulation of the mass balance equations for metabolism as well as alternative statistical approaches to estimating the parameters. Each comparison also evaluated differneces in the metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15; available from @Beck21]. The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection, as a fundamental assumption of metabolic models using *in situ* dissolved oxygen time series from fixed monitoring stations. The prior distributions for EBASE were uninformed using the same large $\rho$ values from @sec-cmpmod and moderate $\mu$ values ($\mu$ = 0.2 $\auni$ for $a$, $\mu$ = 20 $\runi$ for $r$, $\mu$ = 0.251 $\buni$ for $b$).  A one day model optimization period was used for EBASE for comparability with the Odum method.

The common set of metabolic parameters returned by the EBASE and Odum methods are net ecosystem metabolism, gross production, respiration, and gas exchange. Pairwise comparisons of the daily estimates returned by each method were evaluated using simple summary statistics, including Pearson correlation coefficients and linear regression fits to evaluate potential biases in each method. For the latter analysis, differences in intercept estimates less than or greater than zero suggested a bias in magnitude, whereas differences in slope estimates less than or greater than one suggested a bias that depended on relative value of each observation (e.g., bias may differ between lower or higher metabolic estimates). Comparisons of the intercept and slope estimates to zero and one, respectively, were based on confidence intervals from the standard errors of each. Root mean square errors (RMSE) of each linear model were also evaluated. Results were assessed separately for estimates based on the observed or detided dissolved oxygen time series.

<!-- edit -->

@fig-apacmpfig and `r run_reference(id = "tbl-apacmptab")` provide a comparison of the metabolic estimates from the EBASE and Odum methods. Correlations for all estimates, regardless of method or the input time series, were positive and significantly correlated, as expected. The intercept values were all significant, excluding production and respiration with the observed dissolved oxygen, suggesting some level of bias when comparing estimates between the methods. Half of the slope estimates were significantly different for estimates from both the observed and detided dissolved oxygen, with production and gas exchange differing for the observed input and net metabolism and gas exchange differing for the detided input.  The RMSE of each regression comparison were generally lower in the detided results for all comparisons. This latter comparison suggests that detiding the time series produces more comparable estimates between methods.

As noted in the previous paragraph, metabolic estimates between the methods were more similar for results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) as compared to using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for production (P) and positive values for respiration (R), i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., $a$ and $r$ cannot be negative, Equations [-@eq-aprior] and [-@eq-rprior]). Additionally, the effects of tidal advection produced a wider range of estimates for NEM in both methods, whereas the detided results were mostly negative indicating that conclusions about net autotrophy or heterotrophy may be inaccurate using observed dissolved oxygen as input. Overall, detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable (lower standard deviation) and smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

<!-- edit -->

@@

![The default prior distributions for $a$, $r$, and $b$ used in EBASE. All priors are normal distributions and truncated to positive values. $b$ is also truncated to a maximum of 0.504.](figs/priorplot.png){#fig-priorplot fig-align="center"}

@@

![Comparisons of daily metabolic estimates from the EBASE and Odum methods for one year of continuous monitoring data at Apalachicola Bay. Results for net ecosystem metabolism (NEM), gross production (P), respiration (R), and gas exchange (D) are compared using the (a) observed and (b) detided dissolved oxygen time series as input to each model. All values are areal as mmol O$_2$ m$^2$ d$^{-1}$. Black lines show the 1:1 comparison and the red lines show the linear regression fit.](figs/apacmpfig.png){#fig-apacmpfig fig-align="center"}

@@

```{r}
#| label: tbl-ebaseiotab
#| tbl-cap: "EBASE model parameters showing the required input, parameters derived from the model input, and model output.  The model notation for production, respiration, and gas exchange shows both the simple parameter name and the full model component used to estimate the parameter."
ebaseiotab
```

@@

```{r}
#| label: tbl-apacmptab
#| tbl-cap: "Summary statistics of daily metabolic estimates comparing the EBASE and Odum methods for net ecosystem metabolism (NEM), gross primary production (P), respiration (R), and gas exchange (D). Separate comparisons are made for estimates based on observed and detided dissolved oxygen. Summary statistics include Pearson correlation coefficients and intercept, slope, and root mean square error (RMSE) of a linear fit between the same metabolic estimates obtained from each method. Significance of the p-values for the intercept and slope indicates if the estimate is different from zero or one, respectively."
apacmptab
```
